---
title: "Meta-analysis of MetMap Datasets"
author: "Theo Killian"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r, echo = FALSE, results = 'asis'}
# Set global knitr options
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)

# Load required packages
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(stringr)
  library(ggplot2)
  library(ggrepel)
  library(readxl)
  library(writexl)
  library(readr)
  library(grid)
  library(ggvenn)
  library(ggforce)
  library(scales)
  library(FSA)
  library(pheatmap)
  library(plotly)
  library(ggpubr)
  library(umap)
  library(knitr)
  library(RColorBrewer)
  library(DT)
  # library(limma)
  # library(biomaRt)
})

# set project name
project_name <- "metmap_melanoma_analysis"

# Set seed for reproducibility
set.seed(123)
```

# Introduction

This report analyzes metastatic melanoma omic data from MetMap project, as
published in ["A metastasis map of human cancer cell lines" by Jin, et al 2020.](https://www.nature.com/articles/s41586-020-2969-2),
and is described in further detail on the [MetMap website](https://depmap.org/metmap/vis-app/index.html).
The goals of this analysis are:

1) Identify which cell lines have high/low metastatic potential to specific organs.
1) Correlate metastatic potential (CI.95) with penetrance.
1) Compare these scores across organs and cell lines.
1) Highlight patterns of multi-organ metastasis vs. organotropism.

### Some Help Definitions

Here is a short explanation of some of the features that we're about to dive into...

* **Metastatic potential**

DNA barcode abundance detected in each organ relative to the pre-injected population, presented on a log10 scale, range from -4 to 4.

    <= -4: non-metastatic
    -4~-2: (weakly) metastatic, but with low confidence
    >= -2: metastatic, with higher confidence

* **Penetrance**

Percentage of animals that the cell lines were detected via barcode sequencing, ranges 0 to 1

**NOTE: The above definitions were copied verbatim from the MetMap website.**

<!-- Label plots by organ or by tropism (where applicable.) -->

### Methodology To Determine Organ-Specific Tropism

* Establish a baseline penetrance across organs, applying the following formula:

* For each cell line, calculate the **mean penetrance** across all tested organs.

**Formula**

\[
\text{Mean penetrance}_{\text{cell line X}} \;=\;
\frac{\text{penetrance}_{\text{brain}}
      + \text{penetrance}_{\text{lung}}
      + \text{penetrance}_{\text{liver}}
      + \text{penetrance}_{\text{bone}}
      + \text{penetrance}_{\text{kidney}}}{5}
\]

* Calculate organ-specific enrichment (relative penetrance)

* For each *cell line–organ* pair, divide the organ-specific penetrance by the cell line’s mean penetrance.

**Formula**

\[
\text{Relative Penetrance}_{\text{organ}} \;=\;
\frac{\text{penetrance}_{\text{organ}}}
     {\text{Mean penetrance across organs}}
\]

This ratio quantifies how specifically a cell line targets a given organ relative to its overall metastatic capability.

* Threshold setting, defining tropism based on the **relative penetrance**:

| Relative penetrance | Tropism interpretation |
|---------------------|------------------------|
| \(\ge 2.0\)         | **Strong** organ-specific tropism |
| \(\ge 1.5\) and \< 2.0 | **Moderate** organ-specific tropism |
| \< 1.5             | **No clear** organ-specific tropism (generalist or minimal) |

<!-- *Adjust these cut-offs to suit biological context or experimental requirements.* -->

## Read and Process MetMap Data

We read, merge and process the following files:

1) `Supplementary Table 03 MetMap cell line annotation.xlsx`, containing the "metadata" describing all cell lines in this study
1) `Supplementary Table 04 MetMap 500 met potential.xlsx`, organ-specific cell line metastatic scores

These files were downloaded from the [Metmap download page](https://depmap.org/metmap/data/index.html).

```{r eval=FALSE, warning=FALSE, message=FALSE}
read_excel("./data/metmap_data/Supplementary Table 03 MetMap cell line annotation.xlsx") %>%
  dplyr::filter(cancer_type == "melanoma") %>%
  as.data.frame() -> metmap_melanoma_metadata

## get sheet names
sheet_names <- excel_sheets("./data/metmap_data/Supplementary Table 04 MetMap 500 met potential.xlsx")

dplyr::bind_rows(
  ## brain
  read_excel("./data/metmap_data/Supplementary Table 04 MetMap 500 met potential.xlsx",
             sheet = 1) %>%
    dplyr::rename(CCLE_name = names(.)[1]) %>%
    dplyr::mutate(organ = gsub("metp500\\.", "", sheet_names[1])),
  ## lung
  read_excel("./data/metmap_data/Supplementary Table 04 MetMap 500 met potential.xlsx",
             sheet = 2) %>%
    dplyr::rename(CCLE_name = names(.)[1]) %>%
    dplyr::mutate(organ = gsub("metp500\\.", "", sheet_names[2])),
  ## liver
  read_excel("./data/metmap_data/Supplementary Table 04 MetMap 500 met potential.xlsx",
             sheet = 3) %>%
    dplyr::rename(CCLE_name = names(.)[1]) %>%
    dplyr::mutate(organ = gsub("metp500\\.", "", sheet_names[3])),
  ## bone
  read_excel("./data/metmap_data/Supplementary Table 04 MetMap 500 met potential.xlsx",
             sheet = 4) %>%
    dplyr::rename(CCLE_name = names(.)[1]) %>%
    dplyr::mutate(organ = gsub("metp500\\.", "", sheet_names[4])),
  ## kidney
  read_excel("./data/metmap_data/Supplementary Table 04 MetMap 500 met potential.xlsx",
             sheet = 5) %>%
    dplyr::rename(CCLE_name = names(.)[1]) %>%
    dplyr::mutate(organ = gsub("metp500\\.", "", sheet_names[5]))) %>%
  dplyr::filter(CCLE_name %in% metmap_melanoma_metadata$CCLE_name) %>%
  as.data.frame() -> metmap_combined

## calculate z-scores
# metmap_combined %>%
#   dplyr::group_by(organ) %>%
#   dplyr::summarize(threshold_z = mean(penetrance) + sd(penetrance)) %>% 
#   as.data.frame() -> tropic_thresh

## merge metmap numerica and metadata into a single dataframe
metmap_combined %>%
  dplyr::left_join(metmap_melanoma_metadata, by = "CCLE_name") %>%
  # dplyr::left_join(tropic_thresh, by = "organ") %>%
  # dplyr::mutate(
  #   penetrance_z = (penetrance - mean(penetrance, na.rm = TRUE)) / sd(penetrance, na.rm = TRUE),
  #   CI.95_z = (CI.95 - mean(CI.95, na.rm = TRUE)) / sd(CI.95, na.rm = TRUE),
  #   tropism = ifelse(penetrance_z >= threshold_z, "tropic", "metastatic"),
  #   tropism = ifelse(penetrance == 0, "non-metastatic", tropism),
  #   organ_tropism = paste0(organ, "_", tropism)) %>%
  #   group_by(CCLE_name) %>%
  dplyr::mutate(mean_penetrance = mean(penetrance)) %>%
  ungroup() %>%
  dplyr::mutate(
    relative_penetrance = penetrance / mean_penetrance,
    tropism = case_when(
      relative_penetrance >= 2 ~ 'strong_tropism',
      relative_penetrance >= 1.5 & relative_penetrance < 2 ~ 'moderate_tropism',
      TRUE ~ 'no_clear_tropism'),
    tropism = ifelse(penetrance == 0, "no_clear_tropism", tropism),
    organ_tropism = paste0(organ, "_", tropism)) %>%
  dplyr::select(contains("name"), contains("id"), contains("CI."),
                contains("penetrance"), #threshold_z,
                everything()) %>%
  dplyr::select(-cancer_subtype) %>%
  as_tibble() -> metmap_merged

## save the converted metadata table ( for Takis)
metmap_merged %>%
  writexl::write_xlsx(path = "./data/metmap_melanoma_merged_data.xlsx")

## save as RDS
saveRDS(metmap_merged, file = "./data/metmap_melanoma_merged_data.rds")
```

We display the merged and processed MetMap data in a dynamic HTML table below:

```{r}
metmap_merged <- readRDS(file = "./data/metmap_melanoma_merged_data.rds")
## display table
metmap_merged %>%
  dplyr::mutate(across(where(is.numeric), ~ signif(., digits = 3))) %>%
DT::datatable()
```

Breakdown of cell lines by organ tropism:

```{r}
table(metmap_merged$organ_tropism) %>% 
  as.data.frame() %>% 
  dplyr::rename(organ_tropic_class = names(.)[1],
                count = names(.)[2]) %>%
  knitr::kable()
```

<!-- Breakdown of Z-score of Penetrance Threshold -->

<!-- ```{r} -->
<!-- metmap_merged %>%  -->
<!--   dplyr::filter(!duplicated(organ)) %>%  -->
<!--   dplyr::select(organ, threshold_z) %>% -->
<!--   knitr::kable() -->
<!-- ``` -->

<!-- ### Z-Score Formula -->

<!-- Group cell lines by penetrance  -->

<!-- $$ -->
<!-- z_{ij} = \frac{p_{ij} - \mu_i}{\sigma_i} -->
<!-- $$ -->

## Visualizations 

### Heatmap - Metastatic Potential (CI.95)

```{r}
metmap_merged %>% 
  dplyr::select(CCLE_name, CI.95, organ) %>% 
  tidyr::pivot_wider(
    names_from = organ,
    values_from = CI.95) %>%
  tibble::column_to_rownames("CCLE_name") %>% 
  as.data.frame() -> df_meta

metmap_scaled1 <- t(scale(t(df_meta)))  # scale rows, not columns
metmap_scaled1[is.nan(metmap_scaled1)] <- 0

pheatmap(metmap_scaled1,
         fontsize_row = 6,
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         main = "Z-scored CI.95 Metastatic Potential")
```


```{r}
pheatmap(df_meta,
         fontsize_row = 6,
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         main = "Raw CI.95 Metastatic Potential")
```

<!-- * What are "high metastatic potential" cell lines globally, and by organ? -->

<!-- ```{r} -->
<!-- unique(metmap_merged$CCLE_name) -->
<!-- ``` -->

### Heatmap - Penetrance

```{r}
metmap_merged %>% 
  dplyr::select(CCLE_name, penetrance, organ) %>% 
  tidyr::pivot_wider(
    names_from = organ,
    values_from = penetrance) %>%
  tibble::column_to_rownames("CCLE_name") %>% 
  as.data.frame() -> df_pen

metmap_scaled2 <- t(scale(t(df_pen)))  # scale rows, not columns
metmap_scaled2[is.nan(metmap_scaled2)] <- 0

pheatmap(metmap_scaled2,
         fontsize_row = 6,
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         main = "Z-scored Penetrance")
```

```{r}
pheatmap(df_pen,
         fontsize_row = 6,
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         main = "Raw Penetrance Values")
```

<!-- * What are "high penetrance" cell lines globally, and by organ? -->

<!-- ```{r} -->
<!-- unique(metmap_merged$CCLE_name) -->
<!-- ``` -->

## Tropism Classification Distribution

```{r}
metmap_merged %>%
  dplyr::select(CCLE_name, tropism, organ) %>%
  tidyr::pivot_wider(
    names_from = organ,
    values_from = tropism,
    values_fill = NA) %>%
  tidyr::pivot_longer(
    cols = -CCLE_name,
    names_to = "organ",
    values_to = "tropism") %>%
  dplyr::mutate(
    tropism_level = case_when(
      str_detect(tropism, "strong_") ~ "strong tropism",
      str_detect(tropism, "moderate_") ~ "moderate tropism",
      str_detect(tropism, "no_clear_") ~ "no clear tropism",
      TRUE ~ NA_character_),
    fill_color = ifelse(is.na(tropism), "NA", tropism_level)) %>%
  ggplot(aes(x = organ, y = CCLE_name, fill = fill_color)) +
    geom_tile(color = "white") +
    scale_fill_manual(
      values = c(
        "strong tropism" = "orangered",
        "moderate tropism" = "limegreen",
        "no clear tropism" = "steelblue",
        "NA" = "gray90"),
      name = "Tropism") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.text.y = element_text(size = 6),
      panel.grid = element_blank(),
      axis.ticks = element_blank()) +
    labs(
      title = "Organ Tropism by Cell Line",
      x = "Organ Tropism",
      y = "Cell Line")
```

### Violin Plot of Metastatic Potential (CI.95) Across Organs

```{r}
metmap_merged %>%
  ggplot(aes(x = organ, y = CI.95, fill = organ)) +
    geom_violin() +
    geom_jitter(width = 0.2, alpha = 0.5) +
    geom_hline(yintercept = -2, color = "red", linetype = "dashed") +
    theme_minimal() +
    labs(
      title = "Distribution of Metastatic Potential (CI.95) Across Organs",
      x = "Organ",
      y = "CI.95") +
    theme(legend.position = "none")
```

### Box Plot of Metastatic Potential (CI.95) Across Organs

```{r}
metmap_merged %>%
  ggplot(aes(x = organ, y = CI.95, fill = organ)) +
    geom_boxplot() +
    geom_jitter(width = 0.2, alpha = 0.5) +
    geom_hline(yintercept = -2, color = "red", linetype = "dashed") +
    theme_minimal() +
    labs(
      title = "Distribution of Metastatic Potential (CI.95) Across Organs",
      x = "Organ",
      y = "CI.95") +
    theme(legend.position = "none")
```

### Violin Plot of Penetrance Across Organs

```{r}
metmap_merged %>%
  ggplot(aes(x = organ, y = penetrance, fill = organ)) +
    geom_violin() +
    geom_jitter(width = 0.2, alpha = 0.5) +
    theme_minimal() +
    labs(
      title = "Distribution of Penetrance Across Organs",
      x = "Organ",
      y = "Penetrance") +
    theme(legend.position = "none")
```

### Box Plot of Penetrance Across Organs

```{r}
metmap_merged %>%
  ggplot(aes(x = organ, y = penetrance, fill = organ)) +
    geom_boxplot() +
    geom_jitter(width = 0.2, alpha = 0.5) +
    theme_minimal() +
    labs(
      title = "Distribution of Penetrance Across Organs",
      x = "Organ",
      y = "Penetrance") +
    theme(legend.position = "none")
```

### Waterfall Plot - Metastatic Potential (CI.95) by cell line

```{r}
metmap_merged %>% 
  dplyr::arrange(desc(CI.95)) %>%
  dplyr::mutate(rank = row_number()) %>% 
  ggplot(aes(x = rank, y = CI.95, fill = organ)) +
    geom_col() +
    geom_hline(yintercept = -2, color = "red", linetype = "dashed") +
    theme_classic() +
    labs(title = "Waterfall Plot of Metastatic Potential (CI.95)", 
         x = "Ranked Cell Line Samples", 
         y = "CI.95 (Metastatic Potential)", 
         fill = "Organ") +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank()) #-> p1

# plotly::ggplotly(p1)
#p1
```

### Waterfall Plot - Metastatic Potential (CI.95) by "tropic classification"

```{r}
metmap_merged %>% 
  dplyr::arrange(desc(CI.95)) %>%
  dplyr::mutate(rank = row_number()) %>% 
  ggplot(aes(x = rank, y = CI.95, fill = tropism)) +
    geom_col() +
    geom_hline(yintercept = -2, color = "red", linetype = "dashed") +
    theme_classic() +
    labs(title = "Waterfall Plot of Metastatic Potential (CI.95)", 
         x = "Ranked Cell Line Samples", 
         y = "CI.95 (Metastatic Potential)", 
         fill = "Organ") +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank()) #-> p1

# plotly::ggplotly(p1)
#p1
```

### Waterfall Plot - Penetrance by cell line

**NOTE: zero penetrance values have changed to 0.025 so that the organ-specific penetrance can be visualized here.**

```{r}
metmap_merged %>%
  dplyr::mutate(penetrance = if_else(penetrance == 0, 0.025, penetrance)) %>%
  dplyr::arrange(penetrance) %>%
  dplyr::mutate(rank = row_number()) %>% 
  ggplot(aes(x = rank, y = penetrance, fill = organ)) +
    geom_col() +
    theme_classic() +
    labs(title = "Waterfall Plot of Penetrance",
         x = "Ranked Cell Line Samples",
         y = "Penetrance",
         fill = "Organ") +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank()) #-> p2

# plotly::ggplotly(p2)
#p2
```

### Waterfall Plot - Penetrance by "tropic classification"

**NOTE: zero penetrance values have changed to 0.05 so that the organ-specific penetrance can be visualized here.**

```{r}
metmap_merged %>%
  dplyr::mutate(penetrance = if_else(penetrance == 0, 0.025, penetrance)) %>%
  dplyr::arrange(penetrance) %>%
  dplyr::mutate(rank = row_number()) %>% 
  ggplot(aes(x = rank, y = penetrance, fill = tropism)) +
    geom_col() +
    theme_classic() +
    labs(title = "Waterfall Plot of Penetrance",
         x = "Ranked Cell Line Samples",
         y = "Penetrance",
         fill = "Organ") +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank()) #-> p2

# plotly::ggplotly(p2)
#p2
```

### Correlation Between Metastatic Potential and Penetrance

```{r warning=FALSE, message=FALSE}
metmap_merged %>%
  ggplot(aes(x = CI.95, y = penetrance, color = organ)) +
    geom_point(alpha = 0.7) +
    geom_vline(xintercept = -2, color = "red", linetype = "dashed") +
    geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
    # facet_wrap(~ organ) +
    theme_minimal() +
    stat_cor(method = "spearman") +
    labs(
      title = "Correlation between Metastatic Potential (CI.95) and Penetrance",
      x = "Metastatic Potential (CI.95)",
      y = "Penetrance")
```

### Global Correlation Between Metastatic Potential and Penetrance

```{r warning=FALSE, message=FALSE}
metmap_merged %>%
  ggplot(aes(x = CI.95, y = penetrance)) +
    geom_point(alpha = 0.7) +
    geom_vline(xintercept = -2, color = "red", linetype = "dashed") +
    geom_smooth(method = "lm", se = TRUE, linetype = "dashed") +
    # facet_wrap(~ organ) +
    theme_minimal() +
    stat_cor(method = "spearman") +
    labs(
      title = "Correlation between Metastatic Potential (CI.95) and Penetrance",
      x = "Metastatic Potential (CI.95)",
      y = "Penetrance") #-> p4
# plotly::ggplotly(p4)
```

```{r warning=FALSE, message=FALSE}
metmap_merged %>%
  ggplot(aes(x = CI.95, y = penetrance, colour = tropism)) +
    geom_point(alpha = 0.7) +
    geom_vline(xintercept = -2, color = "red", linetype = "dashed") +
    # geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
    # facet_wrap(~ organ) +
    theme_minimal() +
    # stat_cor(method = "spearman") +
    labs(
      title = "Correlation between Metastatic Potential (CI.95) and Penetrance",
      x = "Metastatic Potential (CI.95)",
      y = "Penetrance") #-> p4
# plotly::ggplotly(p4)
```

### Faceted Correlation Between Metastatic Potential and Penetrance

Below we create faceted scatter plots with regression line for each organ. The
vertical red dashed line at `Metastatic Potential = -2` represents where cell
lines are considered "metastatic" or not.

```{r warning=FALSE, message=FALSE}
metmap_merged %>%
  ggplot(aes(x = CI.95, y = penetrance, color = organ)) +
    geom_point(alpha = 0.7) +
    geom_vline(xintercept = -2, color = "red", linetype = "dashed") +
    geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
    facet_wrap(~ organ) +
    theme_minimal() +
    stat_cor(method = "spearman", label.x = -4, label.y = 1.05, size = 3.5) +
    labs(
      title = "Correlation between Metastatic Potential (CI.95) and Penetrance",
      x = "Metastatic Potential (CI.95)",
      y = "Penetrance")
```

### Rose Plots

```{r warning=FALSE, message=FALSE, fig.asp=1.2, fig.width=10, out.width="100%"} 
organ_levels <- c("brain", "lung", "liver", "bone", "kidney")
n_organ <- length(organ_levels)

# Angles at which each petal is centered
angle_centers <- seq(0, 2 * pi - 2 * pi / n_organ, length.out = n_organ)

petal_max_width <- 2 * pi / n_organ # Max width (no more than 1/5 of circle)
petal_min_width <- 0 # petal_max_width * 0.1             # Set a minimum width for aesthetics

metmap_merged %>%
  dplyr::mutate(
    organ = factor(organ, levels = organ_levels),
    organ_index = as.integer(organ) - 1,
    angle_center = angle_centers[organ_index + 1],   # R is 1-based
    penetrance = pmax(penetrance, 0.001),
    petal_width = petal_min_width + (petal_max_width - petal_min_width) * penetrance) %>%
  dplyr::mutate(
    angle_start = angle_center - petal_width / 2,
    angle_end = angle_center + petal_width / 2,
    r0 = 0,
    scaled_r = rescale(CI.95, to = c(0, 1))) %>% 
  ggplot() +
  ggforce::geom_circle(aes(x0 = 0, y0 = 0, r = 1), color = "gray40",
                       linetype = "dashed", inherit.aes = FALSE) +
  geom_arc_bar(
    aes(
      x0 = 0, y0 = 0,
      r0 = r0, r = scaled_r,
      start = angle_start, end = angle_end,
      fill = organ),
    color = "black", alpha = 0.85) +
  coord_fixed() +
  facet_wrap(~CCLE_name, ncol = 5) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 8),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()) +
  labs(
    title = "Rose plots of MetMap melanoma cell lines by metastatized organ",
    fill = "Organ")
```


### Clustering or Organ Tropism Classification

    Test/Approach: Hierarchical clustering or UMAP

    Purpose: Reveal whether certain cell lines cluster by metastatic behavior across organs.

```{r}
umap_result <- umap(metmap_scaled1)
umap_df <- as.data.frame(umap_result$layout)
umap_df$CCLE_name <- rownames(df_meta)
ggplot(umap_df, aes(x = V1, y = V2, label = CCLE_name)) +
  geom_point(alpha = 0.7) +
    geom_text_repel(size = 3, max.overlaps = 50) +
  theme_minimal() +
  labs(title = "UMAP of Cell Lines by Metastatic Potential",
       x = "UMAP1", y = "UMAP2")
```

```{r}
umap_result <- umap(metmap_scaled2)
umap_df <- as.data.frame(umap_result$layout)
umap_df$CCLE_name <- rownames(df_meta)
ggplot(umap_df, aes(x = V1, y = V2, label = CCLE_name)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(size = 3, max.overlaps = 50) +
  theme_minimal() +
  labs(title = "UMAP of Cell Lines by Penetrance",
       x = "UMAP1", y = "UMAP2")
```

```{r}
rownames(metmap_scaled1) <- paste0("CI.95_", rownames(metmap_scaled1))
rownames(metmap_scaled2)   <- paste0("pen_", rownames(metmap_scaled2))
joint_matrix <- cbind(metmap_scaled1, metmap_scaled2)
joint_scaled <- scale(joint_matrix)
umap_result <- umap(joint_scaled)
umap_df <- as.data.frame(umap_result$layout)
umap_df$CCLE_name <- rownames(df_meta)
ggplot(umap_df, aes(x = V1, y = V2, label = CCLE_name)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(size = 3, max.overlaps = 50) +
  theme_minimal() +
  labs(title = "Joint UMAP of Cell Lines by Penetrance and Metastatic Potential",
       x = "UMAP1", y = "UMAP2")
```

### PCA

```{r}
# joint_scaled

# Perform PCA using prcomp()
pca_result <- prcomp(joint_scaled, center = TRUE, scale. = TRUE)

# Summary of variance explained
# summary(pca_result)

# Access principal components
pca_scores <- as.data.frame(pca_result$x)
pca_scores$CCLE_name <- rownames(df_meta)

ggplot(pca_scores, aes(x = PC1, y = PC2, label = CCLE_name)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(size = 3, max.overlaps = 50) +
  theme_minimal() +
  labs(
    title = "PCA of Numeric Data",
    x = "Principal Component 1",
    y = "Principal Component 2"
  )
```

## Statistical Tests

We will run various statistical tests to determine the nature of the relationships between metastatic potential, penetrance and organ in melanoma cell lines. The following is a little background on the statistical tests that we will perform:

1) Kruskal-Wallis Test (Global Non-Parametric Test)

    Purpose: To test whether there is a statistically significant difference in the distribution of a continuous variable (like CI.95 or penetrance) across more than two independent groups (e.g., organs: brain, liver, lung, etc.).

    Why use Kruskal-Wallis instead of ANOVA?

        The variable (e.g., CI.95) is log-transformed and likely non-normally distributed.

        Kruskal-Wallis is a non-parametric test and does not assume normality or equal variance.

    What it tells you: At least one group differs from the others — but not which one(s).

1) Dunn's Test (Post-hoc Pairwise Comparisons)

    Purpose: To identify which specific pairs of groups differ after a Kruskal-Wallis test has found a significant global difference.

    Why not use multiple Wilcoxon tests directly?

    Running many pairwise tests inflates the Type I error rate (false positives).

1) Bonferroni Correction (Control for Multiple Testing)

    Why needed: We're doing multiple pairwise comparisons (e.g., brain vs. liver, brain vs. lung, etc.).

    What it does: Adjusts the p-value threshold to keep the family-wise error rate controlled.

    For example, if you do 10 tests, Bonferroni adjusts the significance threshold from 0.05 to 0.005.

    Effect: More conservative (reduces false positives, may increase false negatives).

### Global Spearman Correlation of Metastatic Potential and Penetrance

```{r warning=FALSE}
# Global correlation
cor_test_global <- cor.test(metmap_merged$CI.95, metmap_merged$penetrance, method = "spearman")
print(cor_test_global)
```

### Per-Organ Spearman Correlation of Metastatic Potential and Penetrance

```{r warning=FALSE}
cor_results_per_organ <- metmap_merged %>%
  dplyr::group_by(organ) %>%
  dplyr::summarize(
    spearman_rho = cor(CI.95, penetrance, method = "spearman"),
    p_value = cor.test(CI.95, penetrance, method = "spearman")$p.value)

print(cor_results_per_organ)
```

### Comparing Metastatic Potential Between Organs

    Test: Kruskal-Wallis test (non-parametric)

    Why: CI.95 is continuous, log-transformed, and likely non-normal. We're comparing distributions across multiple organs.

    Post-hoc: Dunn's test with Bonferroni correction.

```{r warning=FALSE}
# Kruskal-Wallis test for CI.95 across organs
kruskal_result <- kruskal.test(CI.95 ~ organ, data = metmap_merged)
print(kruskal_result)
```

```{r warning=FALSE}
# Dunn's test with Bonferroni correction
dunn_result <- dunnTest(CI.95 ~ organ, data = metmap_merged, method = "bonferroni")
print(dunn_result)
```

### Penetrance Differences Between Organs

    Test: Kruskal-Wallis or ANOVA (depending on distribution)

    Alt: Chi-squared or Fisher's exact test if binarized penetrance (e.g. high ≥ 0.5, low < 0.5).
    
```{r warning=FALSE}
# Kruskal-Wallis test across organs
kruskal_penetrance <- kruskal.test(penetrance ~ organ, data = metmap_merged)
print(kruskal_penetrance)
```

```{r warning=FALSE}
# Dunn's test with Bonferroni correction
dunn_result <- dunnTest(penetrance ~ organ, data = metmap_merged, method = "bonferroni")
print(dunn_result)
```


*Session Info*

```{r sessionInfo}
sessionInfo()
```